Hydra Battles and Cie

### 2021-01-18 doc/chapter-primrec.tex
redaction of Ackermann chapter (continued)

\begin{description}
  \item[zero] the constant function of value $0$ is primitive recursive.
\item[S] The successor function $S:\mathbb{N}\rightarrow\mathbb{N}$ is primitive recursive.
 \item[projections] For any pair $0\leq i<n$, the projection $\pi_{n,i}: \mathbb{N}^n\rightarrow\mathbb{N}$, defined by $\pi_{n,i}(x_0,x_1,\dots,x_{n-1})=x_i$, is primitive recursive.
\item[composition] For any $n$ and $m$, if $h: \mathbb{N}^m\rightarrow\mathbb{N}$, and
$g_0,\dots, g_{m-1}$ are primitive recursive of $n$ arguments, then the function which maps any
tuple $(x_0,\dots,x_{n-1})$ to $h(g_0(x0,\dots,x_{n-1}),\dots, g_{m-1}(x0,\dots,x_{n-1}))$ is primitive recursive.
\item[primitive recursion]
If $g: \mathbb{N}^n\rightarrow\mathbb{N}$ and $h: \mathbb{N}^{n+2}\rightarrow\mathbb{N}$ are primitive recursive, then the function from $\mathbb{N}^{n+1}$ into $\mathbb{N}$ defined by
\begin{align}
f(0,x_1,\dots,x_n)&=g(x_1,\dots,x_n)\\
f(S(p),x_1,\dots,x_n)&=h(p,f(p, x_1,\dots,x_n),  x_1,\dots,x_n)
\end{align} 
is primitive recursive.
\end{description}

@@

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}}
\begin{Coqsrc}
Inductive PrimRec : nat -> Set :=
  | succFunc : PrimRec 1
  | zeroFunc : PrimRec 0
  | projFunc : forall n m : nat, m < n -> PrimRec n
  | composeFunc :
      forall (n m : nat) (g : PrimRecs n m) (h : PrimRec m), PrimRec n
  | primRecFunc :
      forall (n : nat) (g : PrimRec n) (h : PrimRec (S (S n))), 
      PrimRec (S n)
with PrimRecs : nat -> nat -> Set :=
  | PRnil : forall n : nat, PrimRecs n 0
  | PRcons : forall n m : nat, PrimRec n -> PrimRecs n m -> 
                   PrimRecs n (S m).
\end{Coqsrc}

\subsection{A little bit of Semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  from  (\texttt{PrimRec $n$}) to  (\texttt{naryFunc $n$}) (with the helper (\texttt{evalprimRecS $n$ $m$})
from  (\texttt{PrimRecs $n$ $m$}) to  (\texttt{Vector.t (naryFunc $n$) $m$}).

Both functions are mutually defined through dependent pattern matching. We advise the readers who 
would feel uneasy with dependent types to consult Adam Chlipala's \emph{cpdt}  book~\cite{chlipalacpdt2011}. We leave it to the reader  to look at the helper functions in
\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.

@@

 with evalPrimRecs (n m : nat) (fs : PrimRecs n m) {struct fs} :
 Vector.t (naryFunc n) m :=
  match fs in (PrimRecs n m) return (Vector.t (naryFunc n) m) with
  | PRnil a => Vector.nil  (naryFunc a)
  | PRcons a b g gs =>
       Vector.cons _ (evalPrimRec _ g) _  (evalPrimRecs _ _ gs)
  end.
\end{Coqsrc}
Let us take as example the following term\footnote{Of course, we never typed this term \emph{verbatim}, but obtained as the result of some computation we leave it the reader to guess and reproduce.}:

\begin{Coqsrc}
Example weirdPR : PrimRec 1 :=
primRecFunc 0
  (composeFunc 0 1 (PRcons 0 0 zeroFunc (PRnil 0)) succFunc)
  (composeFunc 2 2
    (PRcons 2 1
      (composeFunc 2 1
         (PRcons 2 0 (projFunc 2 1 (le_n 2))
                 (PRnil 2))
         succFunc)
      (PRcons 2 0
        (composeFunc 2 1
          (PRcons 2 0
             (projFunc 2 0
                       (le_S 1 1 (le_n 1)))
             (PRnil 2))
          (projFunc 1 0 (le_n 1))) (PRnil 2)))
    (primRecFunc 1 (composeFunc 1 0 (PRnil 1) zeroFunc)
       (composeFunc 3 2
         (PRcons 3 1
            (projFunc 3 1 (le_S 2 2 (le_n 2)))
            (PRcons 3 0 (projFunc 3 0
                          (le_S 1 2
                                (le_S 1 1 (le_n 1))))
                    (PRnil 3)))
         (primRecFunc 1 (projFunc 1 0 (le_n 1))
                      (composeFunc 3 1
                          (PRcons 3 0
                                  (projFunc 3 1 (le_S 2 2 (le_n 2)))
                                  (PRnil 3))
                          succFunc))))). 
\end{Coqsrc}

@@

>>>>

\begin{description}
  \item[zero] the constant function of value $0$ is primitive recursive.
\item[S] The successor function $S:\mathbb{N}\rightarrow\mathbb{N}$ is primitive recursive.
 \item[projections] For any pair $0< i\leq n$, the projection $\pi_{i,n}: \mathbb{N}^n\rightarrow\mathbb{N}$, defined by $\pi_{i,n}(x_1,x_2,\dots,x_{n})=x_i$, is primitive recursive.
\item[composition] For any $n$ and $m$, if $h: \mathbb{N}^m\rightarrow\mathbb{N}$, and
$g_0,\dots, g_{m-1}$ are primitive recursive of $n$ arguments, then the function which maps any
tuple $(x_0,\dots,x_{n-1})$ to $h(g_0(x0,\dots,x_{n-1}),\dots, g_{m-1}(x0,\dots,x_{n-1}))$ is primitive recursive.
\item[primitive recursion]
If $g: \mathbb{N}^n\rightarrow\mathbb{N}$ and $h: \mathbb{N}^{n+2}\rightarrow\mathbb{N}$ are primitive recursive, then the function from $\mathbb{N}^{n+1}$ into $\mathbb{N}$ defined by
\begin{align}
f(0,x_1,\dots,x_n)&=g(x_1,\dots,x_n)\\
f(S(p),x_1,\dots,x_n)&=h(p,f(p, x_1,\dots,x_n),  x_1,\dots,x_n)
\end{align} 
is primitive recursive.
\end{description}

@@

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.}
\begin{Coqsrc}
Inductive PrimRec : nat -> Set :=
  | succFunc : PrimRec 1
  | zeroFunc : PrimRec 0
  | projFunc : forall n m : nat, m < n -> PrimRec n
  | composeFunc :
      forall (n m : nat) (g : PrimRecs n m) (h : PrimRec m), PrimRec n
  | primRecFunc :
      forall (n : nat) (g : PrimRec n) (h : PrimRec (S (S n))), 
      PrimRec (S n)
with PrimRecs : nat -> nat -> Set :=
  | PRnil : forall n : nat, PrimRecs n 0
  | PRcons : forall n m : nat, PrimRec n -> PrimRecs n m -> 
                   PrimRecs n (S m).
\end{Coqsrc}

\begin{remark}
Beware of the conventions used in the \texttt{primRec} library!
The constructor (\texttt{projFunc $n$ $m$})  is associated with the projection $\pi_{n-m,n}$ and \emph{not}
$\pi_{n, m}$.
For instance, the projection $\pi_{2,5}$ defined by $\pi_{2,5}(a,b,c,d,e)=b$ corresponds to the term
(\texttt{projFunc 5 3 H}), where \texttt{H} is a proof of $3<5$.
 This fact is reported in the comments of \texttt{primRec.v}. We presume that this convention makes it easier to define the evaluation function \texttt{evalProjFunc $n$} (see next sub-section) . Trying the other conention is left as an exercise.
\end{remark}

\subsection{A little bit of Semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  of type
$\texttt{PrimRec}\,n \rightarrow  \texttt{naryFunc}\,n$ (and the helper function 
(\texttt{evalprimRecS $n$ $m$}) of type 
$\texttt{PrimRecs}\,n\,m \rightarrow  \texttt{Vector.t}\,(\texttt{naryFunc}\,n)\,m$.

Both functions are mutually defined through dependent pattern matching. We advise the readers who 
would feel uneasy with dependent types to consult Adam Chlipala's \emph{cpdt}  book~\cite{chlipalacpdt2011}. We leave it to the reader  to look also at the helper functions in
\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.

@@

 with evalPrimRecs (n m : nat) (fs : PrimRecs n m) {struct fs} :
 Vector.t (naryFunc n) m :=
  match fs in (PrimRecs n m) return (Vector.t (naryFunc n) m) with
  | PRnil a => Vector.nil  (naryFunc a)
  | PRcons a b g gs =>
       Vector.cons _ (evalPrimRec _ g) _  (evalPrimRecs _ _ gs)
  end.
\end{Coqsrc}
Let us take as example the following term\footnote{Of course, we never typed this term \emph{verbatim}; we obtained it as the result of some computation we leave it the reader to guess and reproduce.}:

\label{sect:weirdfac}
\begin{Coqsrc}
Example weirdPR : PrimRec 1 :=
primRecFunc 0
  (composeFunc 0 1 (PRcons 0 0 zeroFunc (PRnil 0)) succFunc)
  (composeFunc 2 2
    (PRcons 2 1
      (composeFunc 2 1
         (PRcons 2 0 (projFunc 2 1 (le_n 2))
                 (PRnil 2))
         succFunc)
      (PRcons 2 0
        (composeFunc 2 1
          (PRcons 2 0
             (projFunc 2 0
                       (le_S 1 1 (le_n 1)))
             (PRnil 2))
          (projFunc 1 0 (le_n 1))) (PRnil 2)))
    (primRecFunc 1 (composeFunc 1 0 (PRnil 1) zeroFunc)
       (composeFunc 3 2
         (PRcons 3 1
            (projFunc 3 1 (le_S 2 2 (le_n 2)))
            (PRcons 3 0 (projFunc 3 0
                          (le_S 1 2
                                (le_S 1 1 (le_n 1))))
                    (PRnil 3)))
         (primRecFunc 1 (projFunc 1 0 (le_n 1))
                      (composeFunc 3 1
                          (PRcons 3 0
                                  (projFunc 3 1 (le_S 2 2 (le_n 2)))
                                  (PRnil 3))
                          succFunc))))). 
\end{Coqsrc}

@@

\section{Proving that a given arithmetic function is primitive recursive}

The example in the preceding section clearly shows that, in order to prove that a given arithmetic function
(defined in \gallina{} as usual) is primitive recursive, trying to give  by hand a term  of type \texttt{PrimRec $n$} is not a good method, since such terms may be huge and complex, even for simple arithmetic functions. The method proposed in Library \texttt{primRec} is the following one:

\begin{enumerate}
\item Define a type corresponding to the statement "the function \texttt{$f$:naryFunc $n$} is primitive recursive ''.
\item Prove handy lemmas which may help to prove that a given function is primitive recursive.
\end{enumerate}

Thus, the proof that a function, like \texttt{factorial}, is primitive recursive may be interactive, whithout having to type complex terms at any step of the development.

\subsection{The predicate \texttt{isPR}}

Let $f$ be an arithmetic function of arity $n$. We say that $f$ is primitive recursive if $f$ is \textbf{extensionnaly}
equal to the interpretation of some term of type \texttt{PrimRec $n$}.

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.}
\begin{Coqsrc}
Definition isPR (n : nat) (f : naryFunc n) : Set :=
  {p : PrimRec n | extEqual n (evalPrimRec _ p) f}.  
\end{Coqsrc}

The library \texttt{primRec} contains a large catalogue of lemmas allowing to prove statements 
of the form (\texttt{isPR $n$ $f$}). We won't list all these lemmas here, but give a few examples of
how they may be applied.

\begin{remark}
In the library \texttt{primRec}, all these lemmas are opaque (registered with \texttt{Qed}. Thus they do not allow the user to look at the witness of a proof of a \texttt{isPR} statement. Our example of page\pageref{sect:weirdfac} was built using a  copy of \texttt{primRec.v} where many \texttt{Qed}s have been replaced with
\texttt{Defined}s.
\end{remark}
